A Convolutional Neural Network for Outsole Recognition\tnoteref{t1}

### 2020-04-17 paper/paper-revision.Rnw
Accept edits, minor additional changes

\begin{document}
\maketitle
\begin{abstract}
\svp{Quantitative assessment of class characteristic frequency is a difficult problem, in part because there is currently no way to collect and efficiently process and categorize the quantity of data which would be required for such assessments.
The human labor required to both collect and systematically classify images collected from the population in sufficient quantity to make such assessments has been prohibitive. % I'm not thrilled with this bit, but I do think we need to drive the point home a bit harder as to why we have to automate both parts of this process
In this paper, we discuss a computational method which complements a passive, unmanned scanning device that is currently under development. The automatic class characteristic recognition algorithm described in this paper has been developed with photographs of outsole images, and identifies high-level features} such as circles, stars, and quadrilaterals. \svp{These features are designed to mimic high-level features used by examiners, to facilitate automatic processing of features without resorting to black-box feature sets commonly used in neural networks. The algorithm}  performs well on images with unambiguous shapes and moderate color contrast, though some systematic errors arise from the physical similarities of geometric shapes. Additional improvements may be realized by modifying the labeling scheme to reduce feature overlap and by preprocessing the images to improve contrast. This new approach represents a significant step toward \svp{enabling the collection and application of population-level class characteristic data to support quantitative assessments of the value of class-characteristic based similarity assessments.} % We do NOT want to even hint at automating away anyone's job.
\end{abstract}
%
% \begin{keyword}
% Footwear; class characteristics; \mt{random match probability}; computer vision; neural networks
% \end{keyword}

@@

% \mt{In the United States, forensic analysis of shoe print evidence is varied and difficult.}% this is very imprecise
\svp{Collection and use of footwear evidence varies heavily across jurisdictions within the United States, and, though it is more common for such analyses to be used in Europe, the U.K., and Israel, many of the same fundamental issues remain.}
\svp{It is often difficult to recover footwear evidence with sufficient detail to identify individualizing characteristics, either because of degradation or because the medium did not record the individualizing characteristics in sufficient detail to examine. It is common for prints to be only partially recorded or recovered, which further compounds the difficulty of performing these types of analysis.}
\svp{As a result, }
% \mt{The process of collecting print evidence from a crime scene is not standardized, and it is often impossible to conduct an individualizing analysis (i.e., one based on randomly acquired characteristics) when a print is partial or of low quality. In fact,
the majority of footwear examinations in the United States reach conclusions regarding the correspondence of class characteristics rather than of individualizing characteristics \citep{smith_2011}.
\svp{This tendency is also reflected in the SWGTREAD range of conclusions scale, which has two separate categories to indicate degrees of class characteristic based association out of a range of 7 conclusions.}\fix{add citation}
\svp{Currently, the guidelines also caution against any statistical assessment because accurate data and models to support statistical calculations do not exist.}
\svp{If it were possible to collect a sufficiently large quantity of footwear impression data from a local population, it would be helpful to be able to calculate the frequency of shoes with similar class characteristics in the population in order to characterize the determinative value of a particular piece of evidence.}
% In the cases where individualizing characteristics are not used, it may be both possible and useful to assess the random match probability of the footwear evidence at the level of class characteristics; in other words, given a questioned shoe print, what is the probability that a randomly selected shoe in the relevant population has a similar tread pattern?
Note that this question is distinct from make and model determination since many shoes of different makes and models can still have similar tread patterns \svp{on the class characteristic level.} %\fix{[This is an important point. Should I make a bigger deal out of it throughout the paper?]}% if it fits in, that's a good idea.

% \mt{From a statistical perspective,
\svp{Statistically, } the \svp{coincidental (or random)} match probability is estimated by sampling a large number of shoes from a given population and calculating the observed proportion of shoes in the sample whose tread patterns are similar to the given print.
The primary obstacle to coincidental match probability calculation, then, is one of data collection and annotation.
\svp{The (limited)} number of \svp{population class-characteristic} studies \citep{parent_sandy_significance_nodate,benedict_geographical_2014} \svp{demonstrate the effort required to collect, process, and analyze } even a \svp{moderate amount of data manually}. \svp{However, technological innovations over the past decade have automated previously manual tasks such as license plate reading, tracking individuals on video feeds, and other complex tasks.}
\svp{In these comparable situations, two separate but related technological innovations were required: inexpensive and automatic data collection (e.g. via CCTV or video cameras) hardware, and automatic, accurate data processing software.}
\svp{We are in the process of developing a low-profile scanner which, when embedded in a pedestrian area, passively collects images of shoe treads (a prototype is shown in \autoref{fig:prototype-scanner}).} \svp{In parallel, however, it is necessary to develop software which can process the data from a similarly oriented scanner to generate features which are used to make class characteristic comparisons.} \svp{Through the combination of hardware and software, we hope to enable the collection of population-level footwear data which will facilitate quantitative assessment of class characteristic frequency to support examiner evaluation.}

@@

% From original intro:

Although footwear impressions are more commonly used in practice for footwear analysis than outsole images, our early prototypes of a footwear scanner \svp{suggest} that cameras are more robust than \svp{impression-based systems (e.g.  EverOS) for use in outdoor settings.} \svp{As a result, in this paper describing an initial automated software system for class characteristic recognition, we use outsole images (such as those found on online retail sites), rather than impression-based training data.} \svp{While we recognize that images from the scanner are likely to show some degradation compared to} images found from online retailers, the fundamental features of the \svp{underlying machine learning models are similar.} \svp{In this paper, we present a proof-of-concept model designed to identify class characteristics probabilistically, using labels similar to those used by examiners.} \svp{We discuss the strengths and limitations of both the classification scheme and the neural-network based model, with important implications for future attempts to develop automatic high-level class characteristic recognition systems.}

@@

A reasonable critique of this approach is that images collected from online retailers are of much higher quality than would be expected from the sidewalk scanner proposed in \autoref{sec:Introduction}, and thus the CNN described in this paper is not immediately ready to process real degraded images. \svp{This consideration is justified, as} machine learning methods tend to perform poorly when applied to test data that do not resemble the training data. However, the \svp{use of CNNs to predict class characteristics using the same categories used by human examiners is novel and introduces some interesting complications.} \svp{As a full-featured automatic scanner is still under development, obtaining representative degraded images is not possible (because the scanner's optics are still in flux).} The goal of this work is to evaluate the proposed classification scheme and determine whether CNNs can sufficiently distinguish class characteristic features found on outsoles. Addressing these questions with clean retailer images provides \svp{justification for additional algorithm development as well as continuing investment in optimization of the scanner design}

@@

\svp{In this paper, we have presented} a method \svp{for automatic identification of} geometric class characteristics  \svp{in outsole tread patterns}. A set of geometric class characteristics, defined based on feature sets currently used by examiners, were used to classify a large variety of tread patterns. Labeling these images provided the opportunity to select a manageable number of features and define criteria for labeling these features on newly collected footwear images. Using the convolutional base of VGG16, a pre-trained neural network, we created a new classifier optimized to work with outsole images. This classifier returns the probability of each geometric feature for new outsole images.

% Discussion of weaknesses
Although model predictions are generally accurate, there are a number of systematic misclassifications \svp{and weaknesses} which can be addressed in future iterations of the modeling process. The model has substantial difficulty predicting images with low contrast because image features are obscured. Exploring methods of color correction, such as histogram equalization, may prove useful for eliminating the effect of contrast on predictions. Initial implementations of histogram equalization suggest that the overall accuracy improvement is modest; the model fit to images which have been equalized had approximately 92.5\% validation accuracy, an improvement of 1.5\% over the default model described in this paper.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Quantitative assessment of class characteristic frequency is a difficult problem, in part because there is currently no way to collect and efficiently process the quantity of data that would be required for such assessments without prohibitive amounts of human labor. In this paper, we discuss a computational method which complements a passive, unmanned scanning device that is currently under development. The automatic class characteristic recognition algorithm described in this paper has been developed with photographs of outsole images, and identifies high-level \mt{geometric} features such as circles, stars, and quadrilaterals. These features are designed to mimic high-level features used by examiners to facilitate automatic processing of features without resorting to black-box feature sets commonly used in neural networks. The algorithm performs well on images with unambiguous shapes and moderate color contrast, though some systematic errors arise from the physical similarities of geometric shapes. Additional improvements may be realized by modifying the labeling scheme to reduce feature overlap and by preprocessing the images to improve contrast. This new approach represents a significant step toward enabling the collection and application of population-level class characteristic data to support quantitative assessments of the value of class-characteristic based similarity assessments.
\end{abstract}
%
% \begin{keyword}
% Footwear; class characteristics; \mt{random match probability}; computer vision; neural networks
% \end{keyword}

@@

Collection and use of footwear evidence varies heavily across jurisdictions within the United States, and, though it is more common for such analyses to be used in Europe, the U.K., and Israel, many of the same fundamental issues remain. It is often difficult to recover footwear evidence with sufficient detail to identify individualizing characteristics, either because of degradation or because the medium did not record the individualizing characteristics in sufficient detail to examine. It is also common for prints to be only partially recorded or recovered, which further compounds the difficulty of performing these types of analysis. As a result, the majority of footwear examinations in the United States reach conclusions regarding the correspondence of class characteristics rather than of individualizing characteristics \citep{smith_2011}. This tendency is also reflected in the SWGTREAD range of conclusions scale, which has two separate categories to indicate degrees of class characteristic based association out of a range of 7 conclusions.\fix{add citation} Currently, the guidelines also caution against any statistical assessment because accurate data and models to support statistical calculations do not exist. \mt{However,} if it were possible to collect a sufficiently large quantity of footwear impression data from a local population, it would be helpful to be able to calculate the frequency of shoes \mt{in the population that have similar class characteristics to a given shoe or print in order to characterize the determinative value of that} particular piece of evidence. Note that this question is distinct from make and model determination since many shoes of different makes and models can still have similar tread patterns on the class characteristic level. %\fix{[This is an important point. Should I make a bigger deal out of it throughout the paper?]}% if it fits in, that's a good idea.

Statistically, the coincidental (or random) match probability is estimated by sampling a large number of shoes from a given population and calculating the observed proportion of shoes in the sample whose tread patterns are similar to the given print. The primary obstacle to coincidental match probability calculation, then, is one of data collection and annotation.
The (limited) number of population class-characteristic studies \citep{parent_sandy_significance_nodate,benedict_geographical_2014} demonstrate the effort required to collect, process, and analyze even a moderate amount of data manually. However, technological innovations over the past decade have automated previously manual tasks such as license plate reading, tracking individuals on video feeds, and other complex tasks. In these comparable situations, two separate but related technological innovations were required: inexpensive and automatic data collection (e.g. via CCTV or video cameras) hardware, and automatic, accurate data processing software.
We are in the process of developing a low-profile scanner which, when embedded in a pedestrian area, passively collects images of shoe treads (a prototype is shown in \autoref{fig:prototype-scanner}). In parallel, however, it is necessary to develop software which can process the data from a similarly oriented scanner to generate features which are used to make class characteristic comparisons. Through the combination of hardware and software, we hope to enable the collection of population-level footwear data which will facilitate quantitative assessment of class characteristic frequency to support examiner evaluation.

@@

Although footwear impressions are more commonly used in practice for footwear analysis than outsole images, our early prototypes of a footwear scanner suggest that cameras are more robust than impression-based systems (e.g.  EverOS) for use in outdoor settings. As a result, in this paper describing an initial automated software system for class characteristic recognition, we use outsole images (such as those found on online retail sites), rather than impression-based training data. While we recognize that images from the scanner are likely to show some degradation compared to images found from online retailers, the fundamental features of the underlying machine learning models are similar. In this paper, we present a proof-of-concept model designed to identify class characteristics probabilistically, using labels similar to those used by examiners. We discuss the strengths and limitations of both the classification scheme and the neural-network based model, with important implications for future attempts to develop automatic high-level class characteristic recognition systems.

@@

A reasonable critique of this approach is that images collected from online retailers are of much higher quality than would be expected from the sidewalk scanner proposed in \autoref{sec:Introduction}, and thus the CNN described in this paper is not immediately ready to process real degraded images. This consideration is justified, as machine learning methods tend to perform poorly when applied to test data that do not resemble the training data. However, the use of CNNs to predict class characteristics using the same categories used by human examiners is novel and introduces some interesting complications. As a full-featured automatic scanner is still under development, obtaining representative degraded images is not possible (because the scanner's optics are still in flux). The goal of this work is to evaluate the proposed classification scheme and determine whether CNNs can sufficiently distinguish class characteristic features found on outsoles. Addressing these questions with clean retailer images provides justification for additional algorithm development as well as continuing investment in optimization of the scanner design.

@@

In this paper, we have presented a method for automatic identification of geometric class characteristics in outsole tread patterns. A set of geometric class characteristics were defined based on feature sets currently used by examiners and used to classify a large variety of tread patterns. Labeling these images provided the opportunity to select a manageable number of features and define criteria for labeling these features on newly collected footwear images. Using the convolutional base of VGG16, a pre-trained neural network, we created a new classifier optimized to work with outsole images. This classifier returns the probability of each geometric feature for new outsole images.

% Discussion of weaknesses
Although model predictions are generally accurate, there are a number of systematic misclassifications and weaknesses which can be addressed in future iterations of the modeling process. The model has substantial difficulty predicting images with low contrast because image features are obscured. Exploring methods of color correction, such as histogram equalization, may prove useful for eliminating the effect of contrast on predictions. Initial implementations of histogram equalization suggest that the overall accuracy improvement is modest; the model fit to images which have been equalized had approximately 92.5\% validation accuracy, an improvement of 1.5\% over the default model described in this paper.
